The Impact of Object Immutability on the Java Class Size

### 2020-06-02 paper/article.tex
new format

\documentclass[sigconf]{acmart}
\title{The Impact of Object Immutability on the Java Class Size}
\author{Yegor Bugayenko}{}{}
\email{yegor.bugayenko@huawei.com}
\affiliation{%
  \institution{Huawei Technologies Co., Ltd.}
  \city{Moscow, Russia}
}
\ccsdesc[100]{Software and its engineering~Object oriented languages}
\keywords{Object-Oriented Programming, Immutability, NCSS}
\setcopyright{acmcopyright}
\copyrightyear{2020}
\acmYear{2020}

\usepackage[utf8]{inputenc}
\usepackage[inline]{enumitem}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{verbatimbox}
\usepackage{interval}
\usepackage{hyperref}
\usepackage{minted}
  \setminted{fontsize=\footnotesize}
  \setminted{breaklines}
  \usemintedstyle{bw}
\newcommand{\code}[1]{\texttt{#1}}
\newenvironment{nicetable}
  {\setlength{\parindent}{0em}\medskip\small}
  {\medskip}
\input{total}
\begin{document}

\begin{abstract}
According to subjective opinions of many industry experts,
object immutability is a virtue in object-oriented programming, since it
leads to side-effect free design, cleaner code, better concurrency,
and many other factors. However, it has never been empirically
demonstrated exactly how immutability affects quality metrics of
object-oriented program. In the following research we analyzed
\thetotaljavafiles{} classes from \thetotalrepos{} public Java repositories
to find out how immutability affects the size of code.
\end{abstract}
\maketitle

@@

The only way to modify \texttt{final} attributes is through the constructor
of the class, which in Java by convention has the same name as the class (\text{Book}
in the example above).

@@

To be written...

>>>>

\documentclass[3p,times,procedia]{elsarticle}
\usepackage{ecrc}
\volume{00}
\firstpage{1}
\journalname{Procedia Computer Science}
\runauth{Yegor Bugayenko et al.}
\jid{procs}
\title{The Impact of Object Immutability on the Java Class Size}

\usepackage[bookmarks=false]{hyperref}
  \hypersetup{colorlinks,linkcolor=blue,citecolor=blue,urlcolor=blue}
\usepackage[utf8]{inputenc}
\usepackage[inline]{enumitem}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{verbatimbox}
\usepackage{interval}
\usepackage{hyperref}
\usepackage{minted}
  \setminted{fontsize=\footnotesize}
  \setminted{breaklines}
  \usemintedstyle{bw}
\newcommand{\code}[1]{\texttt{#1}}
\newenvironment{nicetable}
  {\setlength{\parindent}{0em}\medskip\small}
  {\medskip}
\input{total}
\begin{document}
\begin{frontmatter}

\dochead{24th International Conference on Knowledge-Based and Intelligent Information \& Engineering Systems}%
\title{The Impact of Object Immutability on the Java Class Size}
\author[a]{Yegor Bugayenko}
\author[b]{Sergey Zykov}
\address[a]{Huawei Technologies Co., Ltd., Moscow, Russia}
\address[b]{Higher School of Economics, Moscow, Russia}
\begin{abstract}
According to the subjective opinions of many industry experts,
object immutability is a virtue in object-oriented programming, since it
leads to side-effect-free design, cleaner code, better concurrency,
and many other factors. However, it has never been empirically
demonstrated exactly how immutability affects quality metrics of
object-oriented programs. In the following research, we analyzed
\thetotaljavafiles{} classes from \thetotalrepos{} public Java repositories
to find out how immutability affects the size of the code.
\end{abstract}
\begin{keyword}Object-Oriented Programming; Immutability; NCSS\end{keyword}
\end{frontmatter}

@@

The only way to modify \texttt{final} attributes is through the constructor
of the class, which in Java by convention has the same name as the class (\texttt{Book}
in the example above).

@@

The results obtained empirically confirm the hyposesis that immutability
not only leads to better concurrency, side-effect free design, cleaner
contracts between classes, but also to smaller classes, by the amount of
methods and attributes inside.

This is a practical comparison example of two Java libraries designed
by two different groups of programmers for the same purpose: sending emails from Java code.
The first one is \code{commons-email} (version 1.5) by Apache with
a large mutable class \code{SimpleEmail} at the core.
The second one is \code{jcabi-email} (version 1.10) with a set of immutable classes.
Here is how Java source code may look, if it sends an email using \code{commons-email}:

\begin{minted}{text}
Email email = new SimpleEmail();
email.setHostName("smtp.googlemail.com");
email.setSmtpPort(465);
email.setAuthenticator(new DefaultAuthenticator("user", "pwd"));
email.setFrom("yegor256@gmail.com", "Yegor Bugayenko");
email.addTo("friend@jcabi.com");
email.setSubject("How are you?");
email.setMsg("Hi, how are you?");
email.send();
\end{minted}

Here is how the same email sending scenario would be
implemented with \code{jcabi-email}:

\begin{minted}{text}
Postman postman = new Postman.Default(
  new SMTP("smtp.googlemail.com", 465, "user", "pwd")
);
Envelope envelope = new Envelope.MIME(
  new Array<Stamp>(
    new StSender("Yegor Bugayenko <yegor256@gmail.com>"),
    new StRecipient("friend@jcabi.com"),
    new StSubject("How are you?")
  ),
  new Array<Enclosure>(
    new EnPlain("Hi, how are you?")
  )
);
postman.send(envelope);
\end{minted}

In the first example, it is a monster \code{SimpleEmail} class that can do many things, including
sending MIME message via SMTP, creating the message, configuring
its parameters, adding MIME parts to it, and so on. There are 33 private properties, over
a hundred methods, and about two thousands lines of code.

In the second example, there are seven objects instantiated via seven \code{new} calls.
\code{Postman} is responsible for packaging a MIME message; \code{SMTP} is
responsible for sending it via SMTP; stamps (\code{StSender}, \code{StRecipient},
and \code{StSubject}) are responsible for configuring the MIME message before
delivery; enclosure \code{EnPlain} is responsible for creating a MIME part
for the message, which is going to be send. These seven objects being constructed,
encapsulating one into another, and then the postman is asked to \code{send()}
the envelope over the wire.

From a user perspective, there is almost nothing wrong. \code{SimpleEmail} is a powerful
class with multiple controls---a project needs to pick the right one and the job
gets done. However, from a developer perspective \code{SimpleEmail} class is
very difficult to maintain, mostly because the class is large.
Multiple getters and setters, which are the control points of the class, modify
object attributes, configuring its behavior. When a new functionality is
required, a developer has to add new attributes to the class and a new
pair of setters and getters. Of course, such a modification decreases the
cohesion of the class, since there is very little or no interconnection
between newly added attributes and previously existing ones. Every new method
added to such a big class, turns into an isolated island of functionality,
with its own set of attributes.

The immutability of a class makes it difficult to make a class larger
without spending a big amount of efforts for refactoring.
If the \code{SimpleEmail} class would be immutable
in the beginning, it wouldn't be possible to add so many methods into it
and encapsulate so many properties. Because an immutable object only accepts a state
through its constructors. It's difficult to imagine a 33-argument constructor.
When a class is immutable in the first place, its developers are forced
to keep it cohesive and small. Because they can't encapsulate
too much and they can't modify what's encapsulated. Just two or three
arguments of a constructor and the reasonable limit is reached. Everything
on top of that will look strange and clumsy.

The immutable design of \code{jcabi-email} implements exactly the same
email sending functionality, but employs seven classes for that, instead
of one. Of course, the cohesiveness of each of them is much higher than
the one of \code{SimpleEmail}. The length of each of them is below 300
lines of code, which by itself is a perfect indicator of high readability
and maintainability.
Moreover, to extend the functionality of the library, existing classes
don't need to be modified. Each new feature must be added through
creation of new classes and implementing existing interfaces.

Emprical results collected above confirm the logic explained. Larger classes
tend to be mutable, which makes it easier for their authors to make them
even larger when it's necessary.
